Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-01 leadership_across_cultures_contexts.Rmd

In sum, we seek to empirically and systematically accomplish the following four goals: 1) identify the most cross-culturally frequent *qualities of leaders* and the *functions they serve*; 2) identify *features* of the qualities and functions of leaders in reference to theoretical models or novel interpretations; 3) identify the most cross-culturally frequent *costs and benefits* of leadership for both leaders and followers; and 4) explore relationships between measures of leader qualities and functions with group structure and population-level measures. Such exploratory results stand to facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and organization.

@@

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities and functions of leadership that are most and least frequently noted in the ethnographic record.

Our second goal was to discover "features" in our data, e.g., to determine if particular variables tended to covary or cluster together. Such features might lend support to existing theories; for instance they might correspond to "prestige" and "dominance" [@henrich_evolution_2001]. Or they might suggest new and theoretically important dimensions of leadership.

There are numerous approaches to detecting features in matrix data such as ours, many of which employ various dimensionality reduction and clustering techniques. To reduce the dimensionality of the `r length(quality_vars)` leader quality variables and the `r length(function_vars)` leader function variables, all of which take binary values (0, 1), we used logistic PCA from the logisticPCA package. This algorithm projects the natural parameters of the saturated model onto the lower $k-$dimension space that minimizes the Bernoulli deviance [@logisticPCA].

Prior to analysis we removed all rows that had no evidence for any variable, which resulted in `r nrow(pca_data_qualities)` text records for the leader quality variables and `r nrow(pca_data_functions)` text records for the leader function variables. We used cross-validation to optimize rank, $k$, and a tuning parameter, $m$ (the value to approximate the saturated model), per recommendation of the package authors.

@@

We also performed hierarchical cluster analysis on our quality and function variables. We computed distance between our binary vectors using both $1-cor(x, y)$ and asymmetric binary distance (Jaccard distance), and agglomerated using the Ward algorithm. We assessed the robustness of the clusters using the pvclust package [@suzuki_pvclust2006], which computes an unbiased probability that a cluster appears in bootstrapped samples. We used 10,000 bootstrapped samples.

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record.

@@

```{r cluster, fig.cap="Cluster analysis of leader quality variables (left) and leader function variables (right). Distances are $1 - cor$. Ward agglomeration method. p-values computed with 10,000 bootstrap samples using the pvclust package [@pvclust2015]."}
# plot(m_pvclust_qual)
# pvrect(m_pvclust_qual, alpha = 0.9)
knitr::include_graphics("Figures/m_pvclust.both.png")
```

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters. Within each top-level cluster we identify two interpretable and generally strongly supported sub-clusters which are annotated with text. Within the first top-level cluster, we identify a dominant/shaman cluster, which suggests a strong link between dominance-based leadership (i.e., *Authority*, *Aggressiveness*, *Feared*) and shamanism (i.e., *Supernatural*, *Artistic performance*). The second sub-cluster includes qualities of leaders related to material, social, and reproductive success (i.e., *Wealthy*, *Social contacts*, and *Polygnous*). Within the second top-level, we identify a sub-cluster of positive personal characteristics (i.e., *Knowledgeable/intelligent*, *Proper behavior*, *Charisma*, *Fairness*) and a sub-cluster of measures related to social strategies and health (i.e., *Strategic planner*, *Generosity*, *Physically formidable*). The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identify as an organizational functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality*).

>>>>

In sum, we seek to empirically and systematically accomplish the following four goals: 1) identify the most cross-culturally frequent *qualities of leaders* and the *functions they serve*. 2) Identify *features* of the qualities and functions of leaders (e.g., do particular variables tended to covary or cluster together?) in reference to theoretical models or novel interpretations. Such features might lend support to existing theories; for instance they might correspond to "prestige" and "dominance" [@henrich_evolution_2001]. 3) Identify the most cross-culturally frequent *costs and benefits* of leadership for both leaders and followers; and 4) explore relationships between measures of leader qualities and functions with group structure and population-level measures. Such exploratory results stand to facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and organization.

@@

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed confidence intervals for the proportion of text records supporting each variable using generalized linear mixed effects models and random intercepts based on document authors nested within cultures. We computed confidence intervals for the proportion of cultures with evidence for each variable using a cluster bootstrap.

We then explored "features" in our data. There are numerous approaches to detecting features in matrix data such as ours, many of which employ various dimensionality reduction and clustering techniques. We used hierarchical cluster analysis on our quality and function variables. We computed distance between our binary vectors using both $1-cor(x, y)$ and asymmetric binary distance (Jaccard distance), and agglomerated using the Ward algorithm. We assessed the robustness of the clusters using the pvclust package [@suzuki_pvclust2006], which computes an unbiased probability that a cluster appears in bootstrapped samples. We used 10,000 bootstrapped samples.

<!-- To reduce the dimensionality of the `r length(quality_vars)` leader quality variables and the `r length(function_vars)` leader function variables, all of which take binary values (0, 1), we used logistic PCA from the logisticPCA package. This algorithm projects the natural parameters of the saturated model onto the lower $k-$dimension space that minimizes the Bernoulli deviance [@logisticPCA]. Prior to analysis we removed all rows that had no evidence for any variable, which resulted in `r nrow(pca_data_qualities)` text records for the leader quality variables and `r nrow(pca_data_functions)` text records for the leader function variables. We used cross-validation to optimize rank, $k$, and a tuning parameter, $m$ (the value to approximate the saturated model), per recommendation of the package authors. -->

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing 95% CIs as above.

<!-- Our fourth goal was to explore associations between the most common leader qualities and functions, and culture-level variables such as mode of subsistence, social complexity, population density, and group size. We used linear mixed effects models from the lme4 package [@bates2015], with common qualities and functions as outcome variables, culture-level variables as predictors, and documents nested within cultures as grouping variables for random intercepts.  -->

@@

```{r cluster, fig.cap="Cluster analysis of leader quality variables (left) and leader function variables (right). Distances are $1 - cor$. Ward agglomeration method. AU p-values (red) computed with 10,000 bootstrap samples using the pvclust package [@pvclust2015]."}
# plot(m_pvclust_qual)
# pvrect(m_pvclust_qual, alpha = 0.9)
knitr::include_graphics("Figures/m_pvclust.both.png")
```

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters. Within each top-level cluster we identify two interpretable and generally strongly supported sub-clusters which are annotated with text. Within the first top-level cluster, we identify a dominant/shaman cluster, which suggests a strong link between dominance-based leadership (i.e., *Authority*, *Aggressiveness*, *Feared*) and shamanism (i.e., *Supernatural*, *Artistic performance*). The second sub-cluster includes qualities of leaders related to material, social, and reproductive success (i.e., *Wealthy*, *Social contacts*, and *Polygynous*). Within the second top-level, we identify a sub-cluster of positive personal characteristics (i.e., *Knowledgeable/intelligent*, *Proper behavior*, *Charisma*, *Fairness*) and a sub-cluster of measures related to social strategies and health (i.e., *Strategic planner*, *Generosity*, *Physically formidable*).

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identify as an organizational functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality*).

<!-- ## Principle components analyses -->

<!-- A logistic PCA on the `r length(quality_vars)` coded leader quality variables, with rank $k = `r dim(logpca_model_qualities$U)[2]`$ and  $m = `r logpca_model_qualities$m`$ optimized by cross-validation, explained `r signif(logpca_model_qualities$prop_deviance_expl, 3)*100`% of the deviance. We draw on the first two leader quality components for interpretation and additional analyses (see Figure \@ref(fig:qualitiesPCplot)). We identify the first component (PC1) as a *Prestige vs. Dominance* measure. Low values on Leader Qualities PC 1 are strongly associated with the variables *Experienced/accomplished* and *Knowledgeable/intelligence* and moderately associated with *Confidence* and *Bravery*. High values are strongly associated with *Polygynous* and *Authority* and moderately associated with *High quality spouse*. Therefore, the "dominance" side of the *Prestige vs. Dominance* component is also closely liked to reproduction, as demonstrated in the cluster analysis (Figure \@ref(fig:cluster)). Measures of physical formidability and aggressiveness are also moderately associated with the "dominance" side of this component. The second leader qualities component we identify as an *In-group favoritism vs. Social Status* measure. Low values on Leader Qualities PC 2 are strongly associated with *Aggressiveness*, *Bravery*, *Xenophobia*, and *Loyalty*. High Values are strongly associated with *High status* and moderately associated with *Fairness*, *Age*, and *Decisiveness/decision-making*. Therefore, Leader Qualities PC 2 also reflects elements of a prestige-dominance distinction in contrasting aggressiveness with high social status. -->
